Wealth dynamics in the presence of\\network structure and primitive cooperation

### 2019-07-23 manuscript/analysis.tex
Stephen's final wordsmithing.

As mentioned in Section~\ref{giniKindaSucks}, the Gini coefficient is not the ideal measure of inequality for our apocalyptic model. Nonetheless, it is illustrative to see how it varies with respect to the ER $\lambda$ parameter. Figure~\ref{fig:giniVsLambda} depicts the Gini computed \textit{at the onset of stage 3} (before starvation) versus $\lambda$, and confirms that increasing $\lambda$ leads to a decreasing Gini coefficient. Increasing $\lambda$ fosters wealth uniformity through the increased formation of and growth in size of protos. Firstly, a higher percentage of agents join a proto as greater $\lambda$ values lead to fewer isolates in the ER network. As more agents join protos, differences in agent wealth are eliminated as each proto establishes perfect equality amongst its constituent agents, thereby lowering system's overall inequality. Secondly, higher $\lambda$ values lead to larger average proto sizes, as more densely-connected networks increase the likelihood that an agent will join an existing proto rather than form a new one. In much the same way, as smaller protos coalesce into larger ones, the standard inequality between the fragmented protos is eliminated in favor of perfect equality across the larger proto, resulting in a corresponding decrease in the Gini coefficient.

\begin{figure}[hb]
\centering
\includegraphics[width=\columnwidth]{figures/giniVsLambda.png}
\caption{The average Gini coefficient of effective wealth (computed pre-stage 3) for various values of the ER $\lambda$ connectivity parameter, and with both low-noise and high-noise income. 500 agents were used in each simulation. The color band represents a bootstrapped 95\% confidence interval.}
\label{fig:giniVsLambda}
\end{figure}

@@

\subsection{Interpretations, Conjectures and Next Steps}
Unraveling the interplay of structure and dynamics was is a major objective of this offshoot of the Milton and Mudigonda model. The necessarily preliminary analysis reported here shows interesting results in this direction. As the two sources of heterogeneity, both the ambient stochasticity and interaction probability seem to influence the two kinds of inequality indicators.

As expected, figure~\ref{fig:giniVsLambda} shows the role of environmental noise: larger environmental noise produces larger inequality measures. Similarly, figure ~\ref{fig:avgLifespanLambda} shows the role of noise in amplifying differences between isolates and non-isolates: larger the noise, larger the separation between the mean lifetimes of the two populations. Also, as mentioned above, larger $\lambda$ leads to more egalitarian population(figure~\ref{fig:giniVsLambda}); concurrently, larger $\lambda$ leads to larger separation between mean lifetimes of the two populations.

Ideally, we would like to derive these results mathematically, especially the point beyond which the average lifespan of two populations change. Since the SDS has only a few parameters, it would easy to decompose the contributions of the various factors response for mean differential lifetimes as it is unclear whether the differences are due to wealth stabilization induced by proto formation in stage 2, or in stage 3.  Another interesting question is whether the population can be further stratified along proto-size dimensions. We conjecture that protos with larger number of agents will have larger mean lifetimes than protos with smaller numbers in the aggregate. We also conjecture that the time spent in a proto positively influences the mean lifetimes of agents in it.

While the mathematical formulation of the model was presented for a single particle, the system consists of a large ensemble of particles. Many of the characteristics, the differentiation of the system into isolates and non-isolates depend on the order rank order statistics of wealth. Also, More careful statistical tests that go beyond the mean analysis presented here are required to tease out the necessary and sufficient conditions for agents in protos (non-isolates) to consistently outperform the isolates. We are currently pursuing these questions.

>>>>

As mentioned in Section~\ref{giniKindaSucks}, the Gini coefficient is not the ideal measure of inequality for our apocalyptic model. Nonetheless, it is illustrative to see how it varies with respect to the ER $\lambda$ parameter. Figure~\ref{fig:giniVsLambda} depicts the Gini computed \textit{at the onset of Stage 3} (before starvation) versus $\lambda$, and confirms that increasing $\lambda$ leads to a decreasing Gini coefficient. Increasing $\lambda$ fosters wealth uniformity through the increased formation of and growth in size of protos. Firstly, a higher percentage of agents join a proto as greater $\lambda$ values lead to fewer isolates in the ER network. As more agents join protos, differences in agent wealth are eliminated as each proto establishes perfect equality amongst its constituent agents, thereby lowering system's overall inequality. Secondly, higher $\lambda$ values lead to larger average proto sizes, as more densely-connected networks increase the likelihood that an agent will join an existing proto rather than form a new one. In much the same way, as smaller protos coalesce into larger ones, the standard inequality between the fragmented protos is eliminated in favor of perfect equality across the larger proto, resulting in a corresponding decrease in the Gini coefficient.

\begin{figure}[hb]
\centering
\includegraphics[width=\columnwidth]{figures/giniVsLambda.png}
\caption{The average Gini coefficient of effective wealth (computed pre-Stage 3) for various values of the ER $\lambda$ connectivity parameter, and with both low-noise and high-noise income. 500 agents were used in each simulation. The color band represents a bootstrapped 95\% confidence interval.}
\label{fig:giniVsLambda}
\end{figure}

@@

\subsection{Interpretations, Conjectures and Next Steps}
Unraveling the interplay of structure and dynamics is a major objective of this offshoot of the Milton and Mudigonda model. The necessarily preliminary analysis reported here shows interesting results in this direction. As the two sources of heterogeneity, both the ambient stochasticity and interaction probability seem to influence the two kinds of inequality indicators.

As expected, Figure~\ref{fig:giniVsLambda} shows the role of environmental noise: larger environmental noise produces larger inequality measures. Similarly, Figure~\ref{fig:avgLifespanLambda} shows the role of noise in amplifying differences between isolates and non-isolates: the larger the noise, the larger the separation between the mean lifetimes of the two populations. Also, as mentioned above, larger $\lambda$ leads to a more egalitarian population (Figure~\ref{fig:giniVsLambda}); concurrently, larger $\lambda$ leads to larger separation between mean lifetimes of the two populations (Figure~\ref{fig:avgLifespanLambda}).

Ideally, we would like to derive these results mathematically, especially the point beyond which the average lifespan of two populations change. Since the SDS has only a few parameters, it would easy to decompose the contributions of the various factors response for mean differential lifetimes as it is unclear whether the differences are due to wealth stabilization induced by proto formation in Stage 2, or in Stage 3.  Another interesting question is whether the population can be further stratified along proto-size dimensions. We conjecture that protos with larger number of agents will have larger mean lifetimes than protos with smaller numbers in the aggregate. We also conjecture that the time spent in a proto positively influences the mean lifetimes of agents in it.

While the mathematical formulation of the model was presented for a single particle, the system consists of a large ensemble of particles. Many of the characteristics, the differentiation of the system into isolates and non-isolates depend on the order rank order statistics of wealth. Also, a more nuanced statistical analysis that goes beyond the mean analysis presented here is required to tease out the necessary and sufficient conditions for agents in protos (non-isolates) to consistently outperform the isolates. We are currently pursuing these questions.

### 2019-07-23 manuscript/bodycssa19.tex
Stephen's final wordsmithing.

The model presented here was inspired by our search for analogs of the grid world agent-based model (ABM) of Friesen and Mudigonda~\cite{srimil} where foraging agents that pool their resources were shown to on-average outperform non-pooling agents. This model used foraging to mix the population and create opportunities for interaction, and, when certain conditions were met, resource sharing. Our model achieves agent interaction by postulating a static network structure which partially mixes the agents. Of primary interest is whether network effects alone can generate and sustain differences in economic outcomes of otherwise homogeneous agents.

@@

The important role played by social structures in determining economic outcomes of individuals in a society is not in doubt~\cite{granovetter2005, jackson_rev2017}. Still, in the absence of a unifying foundation for sociology and economics, the full impact of this two-way interpenetration of economic and social structures is demonstrated only on a case-by-case basis. The language of social and economic networks affords a first principle integration by simplifying the non-trivial concept of social and economic structure~\cite{martin_lee} to only dyadic (binary) relations among actors\footnote{The concept social structures~\cite{martin_lee} is more general than the widely held belief in computational social science that social structure \textit{is} just a social network.}.

Models of wealth accumulation dynamics in both macro and micro economics typically assume the presence of markets and equilibria. Hence, such models are ill-suited to the study of collective phenomena at an intermediate level. Hence, alternative explanations of aggregate phenomena that match the expressiveness of economic models are required. Analytical sociology~\cite{ch1as_hdbk}, with its emphasis on explanation of collective emergent phenomena using mathematically formulated social mechanisms~\cite{ch2as_hdbk,ch11as_hdbk} anchored at the individual level, is an ideal candidate for this purpose. It is particularly powerful when combined with computational simulation, because when mathematically formulated models reach even a modest level of complexity, they often become analytically intractable. Simulation \textit{in silico} can yield approximate results for these more complex scenarios, which supplement the exact results reached by analytically tractable models.

@@

\label{giniKindaSucks}
The phenomenon under scrutiny is the emergence and evolution of economic inequality in societies with non-trivial social structures that enable economic interactions and cooperation mechanisms. The model we use to answer questions surrounding this phenomenon must possess a dynamic model of wealth accumulation, a model of social structure, and a suitably useful measure of inequality. The dynamics are modeled as Brownian-noise-driven linear dynamical systems, here with constant growth rate; the structure is modeled by networks, here with random graphs; the measures of inequality used are Gini coefficients, and survival time distributions of agents in response to lack of resources. Just as with macroeconomic models, measures of wealth inequality like Gini coefficients are not so well suited for the non-market based wealth dynamics. We discuss this more in section 4.

@@

Apart from enabling interactions between social actors, social structures like institutions also shape the form of cooperation and coordination mechanisms. The institutions can take the form of economic institutions, like banks and cooperatives; or the form of norms, like resource-sharing practices in societies. The Friesen and Mudigonda model~\cite{srimil}consists of a simple resource pooling arrangement where aggregates of agents pool their excess wealth in a common institution called ``proto-institutions,'' agreeing to provide this saved resource to individual agents in times of need. The model of resource pooling used in this work is identical to this model.

@@

In the next section, we discuss the mathematical formulation of the model. In section 3, we present the analysis of our simulation experiments, summarize key findings, and discuss why we chose the \texttt{julia} programming language~\cite{Julia-2017} for our implementation. In section 5, we discuss the limitations of our simple models, extensions to dynamics and networks more expressive than the ones presented here and planned future work.

Before discussing our model in greater mathematical detail, we discuss the model qualitatively, contrasting it with more familiar modeling approaches. The model presented here has much in common with models used in social-reality-inspired models in statistical physics~\cite{redner2001guide}, dynamic process models in network science~\cite{newman2018networks}, computational social science, and agent-based models~\cite{abm_review}; however, our modeling philosophy is somewhere at the interface of agent-based computational economics (ACE)~\cite{tesfatsion} and analytical sociology (AS)~\cite{ch1as_hdbk,ch2as_hdbk}. In ACE, we acknowledge its aspiration to develop bottom-up models of economic systems at all scales but restrain from its enthusiastic use of complex but well-calibrated detailed models of markets and agents~\cite{tesfatsion2017}. In AS, we assent its focus on social mechanisms in explaining social phenomena, but instead rely on simplified mechanisms with few parameters~\cite{ch11as_hdbk} with the specific goal of extracting insights from \textit{stylized} models.

@@

\section{Model}
The model of cooperative wealth accumulation constructed here is best thought of as a stochastic interacting particle system infused with economic sociological semantics. The particle evolves according to a one dimensional diffusion process with constant drift and is driven by a Brownian noise with a boundary condition at the origin corresponding to particle absorption. After crossing a pre-determined threshold in state space, particles above the threshold follow a protocol and coalesce together. In an ensemble of otherwise identical particles, a given particle may coalesce with a subset of other particles. In grid world ABMs, agents interact with other agents by moving around this world. The precise movement protocol encodes how the agents interact (mix) among themselves. In the social network setting, this mixing characteristic is encoded via a graph. Both the conjoined particles and individual particles die upon crossing the origin. This model of interacting diffusing particles can be provided with substantive semantics as follows.

The one-dimensional state space of the particle is identified with the wealth of a social actor (agent). We consider a homogeneous population where all agents are required burn their wealth at a constant specified rate in order to survive. In addition, the agents all gain wealth at a constant rate. The resource draining rate and the resource gaining rate are additive and constitute the drift of the diffusion process. The environmental contingency is modeled by a Brownian noise of a specified intensity. Agents can coalesce to form a cooperative unit deciding to pool their resources and their environmental contingencies into a single unit which we call a \textit{proto-institution} (proto), if they cross a specific wealth threshold. The mixing characteristic of this ensemble is the social network (here ER model).

@@

\noindent where $v$ is the wealth growth rate (the difference of the income and metabolic rate of the agent) and $D$ is the intensity of the Brownian process (white noise process) $w$. $x(t)$ is the state of the particle (wealth) at time $t$. The dynamics can be started at any initial point $x_0 > 0$. Since simulations make use of discrete versions of these equations, a slightly different notation is used\footnote{The discrete time step $\Delta t=1$. $w$ in the discrete setting is a Gaussian distributed random variable $\mathcal{N}(0,\sigma^2) = \mathcal{N}(0,D\Delta t)$ }.

@@

Since the particles are independent, they are independent dynamical systems; proto formation is a higher-level construct imposed on this system and is a constraint on what states are considered viable and which are not. Consider two particles $x_1$ and  $x_2$ that have crossed the threshold for coalescence $x_{thresh}$ ($t_p$ in the next section). After the coalescence event, the state (wealth) of individual particles $x_1(t)$ and  $x_2(t)$ is not relevant for survivability; only the aggregate wealth of the coalescent (proto) determines whether the particles in the coalescent survive. As long as $x_1(t) + x_2(t) >0$, both particles survive. Effectively, the proto is a two-dimensional SDS. Since the wealth of the proto ($p_{12}$) is additive, the aggregate wealth $p_{12}(t) = x_1(t) + x_2(t) $. The aggregate dynamical variable satisfies a SDE\footnote{The result follows from the additivity properties of white noise.}\footnote{The two pictures: the particle perspective and the proto perspective, are equivalent. While it is easier to mathematically analyze the system in the proto perspective, the individual wealth of the particles carries meaning; it is just not useful for studying survival of the proto or the particles within it.}.

@@

As the equations (\ref{e}) and (\ref{s}) show, all agents, irrespective of their starting initial condition and luck, have a non-zero probability of dying. The paper that inspired this work ~\cite{srimil} studied inequality dynamics in the context of extreme scarcity in their foraging world. One can mimic this scarcity by simply turning off the income, starving the agents, and leaving the agents to all inevitably die. The differential rate of death between the isolates and the non-isolates then becomes an important quantity. In the absence of tractable mathematical solutions, simulations experiments which probe this expected differential survival rates help illuminating the role of protos.

We first evolve the system under favorable environmental conditions with steady salary, and identify phases before and after proto formation take place (Stage 1 and 2). After making sure that all surviving agents possess  wealth above the proto formation threshold $x_{thresh}$, we cut off the salary for all agents, starving the agents, leaving them susceptible to the white noise process (Stage 3). In each of the stages, the effective dynamics are different from each other, but the form of the equations remains the same.

To break the homogeneity of this particle ensemble, we propose to let the particles interact with only a random subset of other populations. In other words, we embed the particles on a network (graph); particles interact (form protos) with only their neighbors. In this work, we focus on the simplest \textit{textbook} example of network: Erd\H{o}s-R\'{e}nyi network model~\cite{newman2018networks}. ER models have only one parameter $\lambda$, a parameter that is the average expected degree of any given node that determines the density of connections in the network, and which in this setting determines the expected number of particles available for a particle to coalesce with and form a proto. Commonsense reasoning suggests that a larger $\lambda$ parameter may help increase the average lifetimes of individual agents by enabling greater likelihood of contact with other particles.

Just as equations (\ref{e}) and (\ref{s}) offer insight into the role of dynamics, the key feature of ER models is high probability of formation of giant component at $\lambda>=1$, in which a finite fraction of the population to be connected, have cliques of very high order, and other useful properties guarantee proto formation in large fraction of the population.

As its network properties are well-understood, the network's wealth dynamics allow us to understand more thoroughly the interaction between dynamics and structure, and its role in global behavior. Still, a full analysis of this model remains to be completed and is part of a forthcoming work involving other more expressive network models. As we note subsequently, even this simple setting offers interesting insights regarding the role of network structure in determining important outcomes for the agent.

@@

The simulation~\footnote{All code for this work is available at https://github.com/WheezePuppet/specstar} proceeds as follows. For configurable parameters
$N > 0$,
$\texttt{init\_max} > 0$,
$\texttt{salary} > 0$,
$m > 0$ (metabolic rate),
$t_p > 0$ (the ``proto threshold''),
$\sigma^2 \ge 0$ and
$0 \le \lambda \le 1$:

\begin{enumerate}
\itemsep.1em
\item Create $N$ agents, each with a random initial wealth (uniformly distributed from 0 to \texttt{init\_max}), and related to one another as per a random ER network with parameter $\lambda$.
\item \textbf{Stages 1 and 2\footnote{We define ``Stage 1'' as the period \textit{before} the first proto is formed, ``Stage 2'' as the period \textit{during} which protos are being formed}:} Repeat until all non-isolate agents\footnote{An \textbf{isolate} agent is one with no graph neighbors.} are members of a proto:
    \begin{enumerate}
    \itemsep.1em
    \item Each agent $A$ whose current wealth $\ge t_p$, and who is not currently a member of a proto, chooses at random one of its graph neighbors (call it $B$) whose wealth \textit{also} exceeds $t_p$. (If there are no such neighbors, proceed to the next agent.) If $B$ is already in a proto, have $A$ join $B$'s proto. If not, have $A$ and $B$ form a new proto.
    \item Each agent gains an amount of wealth equal to $(\texttt{salary} - m + \epsilon)$, where $\epsilon \sim \mathcal{N}(0,\sigma^2)$ (white noise).\footnote{Note that this ``gain'' could be negative, in which case the agent, and possibly its proto, may be subject to death exactly as in step (3).}
    \item Each agent \textit{that is in a proto} donates all its wealth in excess of $t_p$ to that proto's balance. (Agents not in a proto maintain their current wealth.)
    \end{enumerate}
    \item \textbf{Stage 3 (starvation):} Repeat until all agents are dead:
        \begin{enumerate}
        \itemsep.1em
        \item Each agent \textit{loses} an amount of wealth equal to $(m + \epsilon), \epsilon \sim \mathcal{N}(0,\sigma^2)$.
        \item If an agent's wealth would drop below zero as a result of this loss, and if it is not a member of a proto, it dies and is removed from the simulation. If it \textit{is} a member of a proto, it withdraws the necessary amount from its proto's balance to remain at zero wealth. If the proto does not have sufficient funds to cover the loss, both the agent and the proto die and are removed from the simulation.
        \end{enumerate}
\end{enumerate}

@@

\section{Analysis}
\input{analysis}
\section{Conclusion and Future Work}
This is one of the first in an ongoing project and the next directions are too many to mention here. Still, we mention a few next steps that adhere to the spirit of our project. These directions all alleviate some of the obvious limitations of the current work.

@@

Even with only a few parameters, the simulation results are hard to visualize and interpret. As the number of parameters increases, as expected in future models, ideas from design and analysis of (computer) experiments may be required. Also, our exploration of the time to death distributions of the population show non-trivial structure. More careful statistical tests that go beyond the mean analysis presented here are required to tease out the necessary and sufficient conditions for agents in protos (non-isolates) to consistently outperform the isolates.

>>>>

The model presented here was inspired by our search for analogs of the grid world agent-based model (ABM) of Friesen and Mudi\-gonda~\cite{srimil} where foraging agents that pool their resources were shown to on-average outperform non-pooling agents. This model used foraging to mix the population and create opportunities for interaction, and, when certain conditions were met, resource sharing. Our model achieves agent interaction by postulating a static network structure which partially mixes the agents. Of primary interest is whether network effects alone can generate and sustain differences in economic outcomes of otherwise homogeneous agents.

@@

The important role played by social structures in determining economic outcomes of individuals in a society is not in doubt~\cite{granovetter2005, jackson_rev2017}. Still, in the absence of a unifying foundation for sociology and economics, the full impact of this two-way interpenetration of economic and social structures is demonstrated only on a case-by-case basis. The language of social and economic networks affords a first principle integration by simplifying the non-trivial concept of social and economic structure~\cite{martin_lee} to only dyadic (binary) relations among actors.\footnote{The ``social structure'' concept~\cite{martin_lee} is more general than the social network specifically, despite the widely held belief in computational social science that they are synonyms.}

Models of wealth accumulation dynamics in both macro and micro economics typically assume the presence of markets and equilibria. Hence, such models are ill-suited to the study of collective phenomena at an intermediate level. Alternative explanations of aggregate phenomena that match the expressiveness of economic models are required. Analytical sociology~\cite{ch1as_hdbk}, with its emphasis on explanation of collective emergent phenomena using mathematically formulated social mechanisms~\cite{ch2as_hdbk,ch11as_hdbk} anchored at the individual level, is an ideal candidate for this purpose. It is particularly powerful when combined with computational simulation, because when mathematically formulated models reach even a modest level of complexity, they often become analytically intractable. Simulation \textit{in silico} can yield approximate results for these more complex scenarios, which supplement the exact results reached by analytically tractable models.

@@

\label{giniKindaSucks}
The phenomenon under scrutiny is the emergence and evolution of economic inequality in societies with non-trivial social structures that enable economic interactions and cooperation mechanisms. The model we use to answer questions surrounding this phenomenon must possess a dynamic model of wealth accumulation, a model of social structure, and a suitably useful measure of inequality. The dynamics are modeled as Brownian-noise-driven linear dynamical systems, here with constant growth rate; the structure is modeled by networks, here with random graphs\footnote{In this paper (as elsewhere) the terms \textbf{graph} and \textbf{network} are exact synonyms.}; the measures of inequality used are Gini coefficients, and survival time distributions of agents in response to lack of resources. Just as with macroeconomic models, measures of wealth inequality like Gini coefficients are not so well suited for the non-market based wealth dynamics. We discuss this more in section 4.

@@

Apart from enabling interactions between social actors, social structures like institutions also shape the form of cooperation and coordination mechanisms. The institutions can take the form of economic institutions, like banks and cooperatives; or the form of norms, like resource-sharing practices in societies. The Friesen and Mudigonda model~\cite{srimil} consists of a simple resource pooling arrangement where aggregates of agents pool their excess wealth in a common institution called a ``proto-institution'' (henceforth, ``\textbf{proto}'') agreeing to provide this saved resource to individual agents in times of need. The model of resource pooling used in this work is identical to this model.

@@

In the next section, we discuss the mathematical formulation of the model. In section 3, we present the analysis of our simulation experiments, summarize key findings, and discuss why we chose the Julia programming language~\cite{Julia-2017} for our implementation. In section 5, we discuss the limitations of our simple models, extensions to dynamics and networks more expressive than the ones presented here and planned future work.

Before discussing our model in greater mathematical detail, we discuss the model qualitatively, contrasting it with more familiar modeling approaches. The model presented here has much in common with models used in social-reality-inspired models in statistical physics~\cite{redner2001guide}, dynamic process models in network science~\cite{newman2018networks}, computational social science, and agent-based models~\cite{abm_review}; however, our modeling philosophy is somewhere at the interface of agent-based computational economics (ACE)~\cite{tesfatsion} and analytical sociology (AS)~\cite{ch1as_hdbk,ch2as_hdbk}. We acknowledge ACE's aspiration to develop bottom-up models of economic systems at all scales, but restrain from its enthusiastic use of complex but well-calibrated detailed models of markets and agents~\cite{tesfatsion2017}. We adhere to AS's focus on social mechanisms in explaining social phenomena, but instead rely on simplified mechanisms with few parameters~\cite{ch11as_hdbk} with the specific goal of extracting insights from \textit{stylized} models.

@@

\section{Model}
The model of cooperative wealth accumulation constructed here is best thought of as a stochastic interacting particle system infused with economic sociological semantics. The particle evolves according to a one-dimensional diffusion process with constant drift and is driven by Brownian noise with a boundary condition at the origin corresponding to particle absorption. After crossing a pre-determined threshold in state space, particles above the threshold follow a protocol and coalesce together. In an ensemble of otherwise identical particles, a given particle may coalesce with a subset of other particles. In grid world ABMs, agents interact with other agents by moving around this world. The precise movement protocol encodes how the agents interact (mix) among themselves. In the social network setting, this mixing characteristic is encoded via a graph. Both the conjoined particles and individual particles die upon crossing the origin. This model of interacting diffusing particles can be provided with substantive semantics as follows.

The one-dimensional state space of the particle is identified with the wealth of a social actor (agent). We consider a homogeneous population where all agents are required burn their wealth at a constant specified rate in order to survive. In addition, the agents all gain wealth at a constant rate. The resource draining rate and the resource gaining rate are additive and constitute the drift of the diffusion process. The environmental contingency is modeled by a Brownian noise of a specified intensity. Agents can coalesce to form a cooperative unit deciding to pool their resources and their environmental contingencies into a single unit which we call a proto-institution (proto), if they cross a specific wealth threshold. The mixing characteristic of this ensemble is the social network (here ER model).

@@

\noindent where $v$ is the wealth growth rate (the difference of the income and metabolic rate of the agent) and $D$ is the intensity of the Brownian process (white noise process) $w$. $x(t)$ is the state of the particle (wealth) at time $t$. The dynamics can be started at any initial point $x_0 > 0$. Since simulations make use of discrete versions of these equations, a slightly different notation is used\footnote{The discrete time step $\Delta t=1$. $w$ in the discrete setting is a Gaussian distributed random variable $\mathcal{N}(0,\sigma^2) = \mathcal{N}(0,D\Delta t)$.}.

@@

Since the particles are independent, they are independent dynamical systems; proto formation is a higher-level construct imposed on this system and is a constraint on what states are considered viable and which are not. Consider two particles $x_1$ and  $x_2$ that have crossed $x_{\textrm{thresh}}$, the threshold for coalescence. After the coalescence event, the state (wealth) of individual particles $x_1(t)$ and  $x_2(t)$ is not relevant for survivability; only the aggregate wealth of the coalescent (proto) determines whether the particles in the coalescent survive. As long as $x_1(t) + x_2(t) >0$, both particles survive. Effectively, the proto is a two-dimensional SDS. Since the wealth of the proto ($p_{12}$) is additive, the aggregate wealth $p_{12}(t) = x_1(t) + x_2(t) $. The aggregate dynamical variable satisfies a SDE\footnote{The result follows from the additivity properties of white noise.}\footnote{The two pictures: the particle perspective and the proto perspective, are equivalent. While it is easier to mathematically analyze the system in the proto perspective, the individual wealth of the particles carries meaning; it is just not useful for studying survival of the proto or the particles within it.}.

@@

As the equations (\ref{e}) and (\ref{s}) show, all agents, irrespective of their starting initial condition and luck, have a non-zero probability of dying. The paper that inspired this work ~\cite{srimil} studied inequality dynamics in the context of extreme scarcity in their foraging world. One can mimic this scarcity by simply ``turning off'' the income at some point, starving the agents, and leaving the agents to all inevitably die. The differential rate of death between the isolates and the non-isolates then becomes an important quantity. In the absence of tractable mathematical solutions, simulation experiments which probe this expected difference in survival rates help illuminate the role of protos.

We first evolve the system under favorable environmental conditions with steady salary, during what we define as ``Stages 1 and 2.'' \textbf{Stage 1} is the phase of the dynamic model \textit{before} which any agents have formed protos (since their wealth has not yet reached $x_{\textrm{thresh}}$), and \textbf{Stage 2} is the phase \textit{during which} agents are forming protos. Once all non-isolate agents have joined a proto, \textbf{Stage 3} commences, at which point we cut off the salary for all agents, starving them and leaving them susceptible to the white noise process. The effective dynamics are different in each stage, but the form of the equations remains the same.

To break the homogeneity of this particle ensemble, we propose to let the particles interact with only a random subset of other populations. In other words, we embed the particles on a network (graph); particles interact (form protos) with only their neighbors. In this work, we focus on the simplest \textit{textbook} example of network: Erd\H{o}s-R\'{e}nyi network model~\cite{newman2018networks}. ER models have only one parameter ($\lambda$), which is the \textit{average expected degree} of any given node. This determines the density of connections in the network, and in this setting determines the expected number of particles available for a particle to coalesce with and form a proto. Commonsense reasoning suggests that a larger $\lambda$ parameter may help increase the average lifetimes of individual agents by enabling greater likelihood of contact with other particles.

Just as equations (\ref{e}) and (\ref{s}) offer insight into the role of dynamics, the key feature of ER models is the high probability of the formation of a ``giant component'' at $\lambda \ge 1$. When a giant component exists, a finite fraction of the population is connected. Networks with this property also tend to have cliques of very high order, and other useful properties that bear upon proto formation.

As its network properties are well-understood, the network's wealth dynamics allow us to understand more thoroughly the interaction between dynamics and structure, and their role in global behavior. Still, a full analysis of this model remains to be completed and is part of a forthcoming work involving other more expressive network models. As we note subsequently, even this simple setting offers interesting insights regarding the role of network structure in determining important outcomes for the agent.

@@

The simulation\footnote{All code for this work is available at \texttt{https://github.com/WheezePuppet/specstar}.} proceeds as follows. For configurable parameters
$N > 0$,
$\texttt{init\_max} > 0$,
$\texttt{salary} > 0$,
$m > 0$ (metabolic rate),
$x_{\textrm{thresh}} > 0$ (the ``proto threshold''),
$\sigma^2 \ge 0$ and
$0 \le \lambda \le 1$:

\begin{enumerate}
\itemsep.1em
\item Create $N$ agents, each with a random initial wealth (uniformly distributed from 0 to \texttt{init\_max}), and related to one another as per a random ER network with parameter $\lambda$.
\item \textbf{Stages 1 and 2:} Repeat until all non-isolate agents\footnote{An \textbf{isolate} agent is one with no graph neighbors.} are members of a proto:
    \begin{enumerate}
    \itemsep.1em
    \item Each agent $A$ whose current wealth $\ge x_{\textrm{thresh}}$, and who is not currently a member of a proto, chooses at random one of its graph neighbors (call it $B$) whose wealth \textit{also} exceeds $x_{\textrm{thresh}}$. (If there are no such neighbors, proceed to the next agent.) If $B$ is already in a proto, have $A$ join $B$'s proto. If not, have $A$ and $B$ form a new proto.
    \item Each agent gains an amount of wealth equal to $(\texttt{salary} - m + \epsilon)$, where $\epsilon \sim \mathcal{N}(0,\sigma^2)$ (white noise).\footnote{Note that this ``gain'' could be negative, in which case the agent, and possibly its proto, may be subject to death exactly as in step (3).}
    \item Each agent \textit{that is in a proto} donates all its wealth in excess of $x_{\textrm{thresh}}$ to that proto's balance. (Agents not in a proto maintain their current wealth.)
    \end{enumerate}
    \item \textbf{Stage 3 (starvation):} Repeat until all agents are dead:
        \begin{enumerate}
        \itemsep.1em
        \item Each agent \textit{loses} an amount of wealth equal to $(m + \epsilon), \epsilon \sim \mathcal{N}(0,\sigma^2)$.
        \item If an agent's wealth would drop below zero as a result of this loss, and if it is not a member of a proto, it dies and is removed from the simulation. If it \textit{is} a member of a proto, it withdraws the necessary amount from its proto's balance to remain at zero wealth. If the proto does not have sufficient funds to cover the loss, both the agent and the proto die and are removed from the simulation.
        \end{enumerate}
\end{enumerate}

@@

\section{Analysis}
\input{analysis}
\section{Conclusion and Future Work}
This is one of the first phases of an ongoing project, and the next directions are too many to enumerate here. Still, we mention a few next steps that adhere to the spirit of our project. These directions all alleviate some of the obvious limitations of the current work.

@@

Even with only a few parameters, the simulation results are challenging to visualize and interpret. As the number of parameters increases, as expected in future models, ideas from design and analysis of (computer) experiments may be required. Also, our exploration of the time to death distributions of the population show non-trivial structure. More careful statistical tests that go beyond the mean analysis presented here are required to tease out the necessary and sufficient conditions for agents in protos (non-isolates) to consistently outperform the isolates.

### 2019-07-23 manuscript/verification.tex
Stephen's final wordsmithing.

For a sensible range of parameter settings, the life history of a single
simulation run follows an expected pattern. Assuming a positive growth rate
(\textit{i.e.}, salary > metabolic rate), each agent's wealth rises unsteadily
during stage 1, until eventually the first pair of neighboring agents who each
reach the threshold form a proto-institution. Throughout stage 2, these agents
contribute all wealth in excess of the threshold to that proto, and so their
proto's balance rises unsteadily while their personal wealth remains at the
constant threshold. Meanwhile, the other agents also reach the threshold at
various points in time, and also form or join protos, until every non-isolate
(that is, every node with at least one neighbor) is a member of a proto. Stage
3 (the starvation period) then commences, with isolates drawing on a greater
personal wealth than the non-isolates. When an isolate reaches zero wealth, it
dies; when a non-isolate reaches zero, it draws from its (shared) proto balance
until that too, reaches zero, and it dies.

@@

This history can be seen in Figure~\ref{fig:singleRun}. The top plot depicts
agent wealth at every iteration, and the bottom plot shows the balances of the
protos at the same points in time. (No protos exist until stage 2, by
definition.) Note that the isolates (orange lines) never form protos, and
therefore begin the starvation stage with a higher personal wealth to draw
from.

@@

We verified that changes to basic parameters all have the expected effect:
lowering the initial wealth delays the onset of stage 2; a higher $\sigma^2$
for the income distribution makes the lines less jagged; a higher metabolic
rate hastens extinction; \textit{etc.}

@@

    The Gini coefficient of the agent population as seen in the top plot of Figure~\ref{fig:giniHistory} is influenced by two distinct dynamics: loss/accumulation of wealth and the formation/death of proto-institutions. Over the course of Stage 1 and the beginning of Stage 3, this change in agent wealth is exclusively responsible for changes in the Gini coefficient. As agents accumulate wealth over Stage 1 and 2, the size of wealth differentials shrinks relative to absolute agent wealth, leading to the declining Gini coefficient. The opposite effect occurs during Stage 3 as agent starvation leads to the relative growth of these wealth differentials. The variability in starvation rates further contributes to the increasing Gini coefficient.

>>>>

For a sensible range of parameter settings, the life history of a single
simulation run follows an expected pattern. Assuming a positive growth rate
(\textit{i.e.}, salary > metabolic rate), each agent's wealth rises unsteadily
during Stage 1, until eventually the first pair of neighboring agents who each
reach the threshold form a proto-institution. Throughout Stage 2, these agents
contribute all wealth in excess of the threshold to that proto, and so their
proto's balance rises unsteadily while their personal wealth remains at the
constant threshold. Meanwhile, the other agents also reach the threshold at
various points in time, and also form or join protos, until every non-isolate
(that is, every node with at least one neighbor) is a member of a proto. Stage
3 (the starvation period) then commences, with isolates drawing on a greater
personal wealth than the non-isolates. When an isolate reaches zero wealth, it
dies; when a non-isolate reaches zero, it draws from its (shared) proto balance
until that too, reaches zero, and it dies.

@@

This history can be seen in Figure~\ref{fig:singleRun}. The top plot depicts
agent wealth at every iteration, and the bottom plot shows the balances of the
protos at the same points in time. (No protos exist until Stage 2, by
definition.) Note that the isolates (orange lines) never form protos, and
therefore begin the starvation stage with a higher personal wealth to draw
from.

@@

We verified that changes to basic parameters all have the expected effect:
lowering the initial wealth delays the onset of Stage 2; a higher $\sigma^2$
for the income distribution makes the lines less jagged; a higher metabolic
rate hastens extinction; \textit{etc.}

@@

    The Gini coefficient of the agent population as seen in the top plot of Figure~\ref{fig:giniHistory} is influenced by two distinct dynamics: loss/accumulation of wealth and the formation/death of proto-institutions. Over the course of Stage 1 and the beginning of Stage 3, this change in agent wealth is exclusively responsible for changes in the Gini coefficient. As agents accumulate wealth over Stage 1 and 2, the size of wealth differentials shrinks relative to absolute agent wealth, leading to the declining Gini coefficient. The opposite effect occurs during Stage 3 as agent starvation leads to the relative growth of these wealth differentials. The variability in starvation rates further stimulates the increasing Gini coefficient.
